---
title: "How does preference for residing along Charles River influences Boston housing prices "
author: "Jiayi Zheng"
date: "04/20/2022"
output: pdf_document
abstract: This paper revisits the 1978 Boston housing market data used in [@A1]. The paper first explores the relationship of Boston housing prices with many other variables, such as per capita crime rate by town, average number of rooms per dwelling, lower status of the population, etc., with emphasis on showing people's preference for residing along river tract. Then, the paper builds a suitable Boston housing price prediction model using suitable predictors based on analysis results.
bibliography: reference.bib
---
Website for github^[https://github.com/wholesomefcas/BostonHousing]

```{r, echo=FALSE, include=FALSE, message=FALSE}
library(readxl)
library(tidyverse)
library(janitor)
library(tinytex)
library(tesseract)
library(reader)
library(corrplot)
library(broom)
library(car) # The following object is masked from ‘package:purrr’:
library(lars)
library(randomForest)
library(Metrics)
```

```{r, echo=FALSE, include=FALSE, message=FALSE}
#cite the packages
citation("tidyverse")
citation("readxl")
citation("janitor")
print(citation("tinytex"), bibtex=TRUE)
citation("dplyr")
citation("ggplot2")
citation()
```


# 1 Introduction

Residing along river tracts offers many benefits. One could easily go out for a jogging with the company of pleasant river sounds. However, expressing the effect of residing along river tracts in monetary terms is not so easy, although it is generally expected that such effect has to be a positive one. One of the reasons for this is that specifying the functional form of the relationship between house prices and whether a house is located along a river tract can be hard. The common assumption of a linear relationship may cause a over-fitting problem, should the actual relationship is seriously non-linear. This paper attempts to solve the concern over function form specification by using machine learning, specifically, LASSO regression technique, and compare its performance with Linear regression model on the basis of root-mean-square error (RMSE).

This paper first explores the relationship of Boston housing prices with many other variables, such as per capita crime rate by town, average number of rooms per dwelling, lower status of the population, etc., with emphasis on showing people's preference for residing along river tract. Then, the paper builds a suitable Boston housing price prediction model using suitable predictors based on analysis results.

```{r, echo=FALSE, message=FALSE, include=FALSE}
boston_orignial <- read.table("boston.txt", sep = ' ',fill = TRUE, 
                              header = FALSE)

boston <- boston_orignial

# put all NA values to the end without changing the order of non-NA values
reorder_na <- function(x){
  na <- is.na(x)
  c(x[!na], x[na])
}

# apply reorder_na to the data
for (j in 1:(nrow(boston_orignial))){
    boston[j,] <- reorder_na(boston[j,])
}

# delete unneeded columns
boston <- boston[,-c(15:25)]

# attach the entries in every other row to the end of the the previous row
for (i in 1:(82/2)){
  for (j in 1:3){
    boston[2*i-1,11+j] <- boston[2*i,j]
  }
}

for (i in c(83, 86,89)){
  boston[i,11] <- boston[i+1,1]
  boston[i,12] <- boston[i+2,1]
  boston[i,13] <- boston[i+2,2]
  boston[i,14] <- boston[i+2,3]
}

for (i in 0:((nrow(boston)-92-1)/2)){
  for (j in 1:3){
    boston[92+i*2,11+j] <- boston[92+i*2+1,j]
  }
}


# delete all unneeded rows
boston <- boston[-c(seq(2,83,by=2),84,85,87,88,90,91,seq(93,nrow(boston),by=2)),]

sum(is.na(boston)) # no missing values


colnames(boston) = c("CRIM","ZN","INDUS","CHAS","NOX",
              "RM","AGE","DIS","RAD","TAX",
              "PTRATIO","B","LSTAT","MEDV")

#boston$CHAS <- ifelse(boston$CHAS == 1, "tract bounds river")
```

# 2 Data 

The data is analysis and cleaned in R [@R1] using readxl [@P2], jenitor [@P3], the tidyverse [@P1], and dplyr packages [@P5] in the Rstudio.The paper is knitted in Rmarkdown.\

## 2.1 Data source 
The data is downloaded from [@data]. The data comes from [@A1]. 
 
## 2.2 Data overview and cleaning 

There are 508 observations for 14 variables. There are no missing values in this data set. The first 21 lines are removed, because those lines only contain explanatory notes but no actual data.\

The variables in order are [@data]:\
 
 CRIM:     per capita crime rate by town.\
 ZN:       proportion of residential land zoned for lots over 25,000 sq.ft\
 INDUS:    proportion of non-retail business acres per town\
 CHAS:     Charles River dummy variable (= 1 if tract bounds river; 0 otherwise)\
 NOX:      nitric oxides concentration (parts per 10 million)\
 RM:       average number of rooms per dwelling\
 AGE:      proportion of owner-occupied units built prior to 1940\
 DIS:      weighted distances to five Boston employment centres\
 RAD:      index of accessibility to radial highways\
 TAX:      full-value property-tax rate per $10,000\
 PTRATIO:  pupil-teacher ratio by town\
 B:        1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town\
 LSTAT:    % lower status of the population\
 MEDV:     Median value of owner-occupied homes in $1000's\

## 2.3 Data summary 
```{r, echo=FALSE, message=FALSE}
knitr::kable(summary(boston)[,1:7], caption = "Boston Housing Data Descriptive Statistics Table Part 1")
```

```{r, echo=FALSE, message=FALSE}
knitr::kable(summary(boston)[,8:14], caption = "Boston Housing Data Descriptive Statistics Table Part 2")
```

The summary of `MEDV` shows the minimum median value of owner-occupied homes is \$5k and the maximum median value of owner-occupied homes is \$50k. The median of `MEDV` is about \$21.2k.\

# 3 Analysis

## 3.1 Distribution of the response variable `MEDV`

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=boston, aes(x=MEDV))+
  geom_histogram(fill='blue', color='black', binwidth=3)+
  labs(x='MEDV', title=" Figure 1: Histogram of MEDV")
```

The dependent variable, `MEDV`, are a little skewed to the right. Because the skewness is not much, we do not need to do a transformation by Log to correct it.

## 3.2 Whether locating along Charles River (= 1 if tract bounds river; 0 otherwise) 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
boston %>%
ggplot() +
geom_histogram(aes(x = MEDV, fill = factor(CHAS)), position ="stack", bins = 15) +
facet_wrap(~factor(CHAS), nrow = 1) +
theme_minimal() +
labs(caption = "Figure 2: MEDV in different CHAS (CHAS = 1 if tract bounds river; 0 otherwise)")
```
Grouping by whether locating along Charles River (= 1 if tract bounds river; 0 otherwise), it can be seen that most houses in our data does not locate along Charles River. For those that does not locate along Charles River, the distribution of `MEDV` is about symmetric. For those that does locate along Charles River, the distribution of `MEDV` is about uniform. The 2 ranges are similar.\ 

## 3.3  Accessibility to radial highways `RAD`
```{r, echo=FALSE, message=FALSE, warning=FALSE}
boston %>%
ggplot() +
geom_histogram(aes(x = MEDV, fill = factor(CHAS)), position ="dodge", bins = 15) +
facet_wrap(~RAD, nrow = 3) +
theme_classic() +
labs(caption = "Figure 3: Median value of owner-occupied homes in $1000's for different accessibility to radial highways")
```
Grouping by accessibility to radial highways and indicated by whether locating along Charles River, it can be seen that most houses in our data has accessibility index 4, 5, or 24. For those that does not locate along Charles River, the distributions of `MEDV` are about symmetric for those with accessibility index 1, 3, 6, 7, or 8. For those that does not locate along Charles River, the distributions of `MEDV` are right-skewed for those with accessibility index 2, 4, 5, or 24. For those that does locate along Charles River, the distributions of `MEDV` are about uniform for different accessibility indexes. It seems that accessibility to radial highways does not influence the distribution of `MEDV` by much, whether houses are located along Charles River or not.\ 

## 3.4 Per capita crime rate by town
```{r, echo=FALSE, message=FALSE, warning=FALSE}
boston %>%
ggplot() +
geom_point(aes(x = MEDV, y = CRIM, color = factor(CHAS))) +
theme_minimal() +
labs(caption = "Figure 4: Median value of owner-occupied homes in $1000's for different per capita crime rate by town")
```
Plotting Per capita crime rate by town vs `MEDV`, grouped by whether locating along Charles River, it can be seen that higher crime rate decreases house prices, whether houses are located along Charles River or not. 

## 3.5 Proportion of residential land zoned for lots over 25,000 sq.ft.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
boston %>%
ggplot() +
geom_point(aes(x = MEDV, y = ZN, color = factor(CHAS))) +
theme_minimal() +
labs(caption = "Figure 5: Median value of owner-occupied homes for different proportion of residential land zoned for lots")
```

Plotting Proportion of residential land zoned for lots over 25,000 sq.ft. vs `MEDV`, grouped by whether locating along Charles River, it can be seen that higher Proportion of residential land zoned for lots increases house prices, whether houses are located along Charles River or not. However, for the same level of proportion of residential land zoned for lots, average `MEDV` seems to be always higher for houses located along Charles River. 

## 3.6 Proportion of non-retail business acres per town
```{r, echo=FALSE, message=FALSE, warning=FALSE}
boston %>%
ggplot() +
geom_point(aes(x = MEDV, y = INDUS, color = factor(CHAS))) +
theme_minimal() +
labs(caption = "Figure 6: Median value of owner-occupied homes for different proportion of non-retail business acres per town")
```

Plotting proportion of non-retail business acres per town vs `MEDV`, grouped by whether locating along Charles River, it can be seen that higher proportion of non-retail business acres per town decreases house prices, whether houses are located along Charles River or not. For the same level of proportion of non-retail business acres per town, average `MEDV` seems to be about the same for houses located along Charles River and for houses not located along Charles River.

## 3.7 Nitric oxides concentration (parts per 10 million)
```{r, echo=FALSE, message=FALSE, warning=FALSE}
boston %>%
ggplot() +
geom_point(aes(x = MEDV, y = NOX, color = factor(CHAS))) +
theme_minimal() +
labs(caption = "Figure 7: Median value of owner-occupied homes for different nitric oxides concentration")
```

Plotting Nitric oxides concentration (parts per 10 million) vs `MEDV`, grouped by whether locating along Charles River, it can be seen that higher Nitric oxides concentration (parts per 10 million) decreases house prices, whether houses are located along Charles River or not. For the same level of Nitric oxides concentration (parts per 10 million), average `MEDV` seems to be about the same for houses located along Charles River and for houses not located along Charles River.

## 3.8 Average number of rooms per dwelling
```{r, echo=FALSE, message=FALSE, warning=FALSE}
boston %>%
ggplot() +
geom_point(aes(x = MEDV, y = RM, color = factor(CHAS))) +
theme_minimal() +
labs(caption = "Figure 8: Median value of owner-occupied homes for different average number of rooms per dwelling")
```
Plotting average number of rooms per dwelling vs `MEDV`, grouped by whether locating along Charles River, it can be seen that higher average number of rooms per dwelling increases house prices in a near linear way, whether houses are located along Charles River or not. The positive linear relationship between average number of rooms per dwelling and `MEDV` seems to be the same for houses located along Charles River and for houses not located along Charles River.

## 3.9 Proportion of owner-occupied units built prior to 1940
```{r, echo=FALSE, message=FALSE, warning=FALSE}
boston %>%
ggplot() +
geom_point(aes(x = MEDV, y = AGE, color = factor(CHAS))) +
theme_minimal() + 
labs(caption = "Figure 9: Median value of owner-occupied homes for different proportion of owner-occupied units built prior to 1940")
```
Plotting proportion of owner-occupied units built prior to 1940 vs `MEDV`, grouped by whether locating along Charles River, it can be seen that higher proportion of owner-occupied units built prior to 1940 decreases house prices, whether houses are located along Charles River or not. For the same level of proportion of owner-occupied units built prior to 1940, average `MEDV` seems to be higher for houses located along Charles River than for houses not located along Charles River.

## 3.10 Weighted distances to five Boston employment centres
```{r, echo=FALSE, message=FALSE, warning=FALSE}
boston %>%
ggplot() +
geom_point(aes(x = MEDV, y = DIS, color = factor(CHAS))) +
theme_classic() + 
labs(caption = "Figure 10: Median value of owner-occupied homes for different weighted distances to five Boston employment centres")
```
Plotting weighted distances to five Boston employment centres vs `MEDV`, grouped by whether locating along Charles River, it can be seen that higher weighted distances to five Boston employment centres increases house prices in a near linear way, whether houses are located along Charles River or not. The positive linear relationship between weighted distances to five Boston employment centres and `MEDV` seems to be the same for houses located along Charles River and for houses not located along Charles River.

## 3.11 Pupil-teacher ratio by town
```{r, echo=FALSE, message=FALSE, warning=FALSE}
boston %>%
ggplot() +
geom_point(aes(x = MEDV, y = PTRATIO, color = factor(CHAS))) +
theme_classic() + 
labs(caption = "Figure 11: Median value of owner-occupied homes for different pupil-teacher ratio by town")
```

Plotting pupil-teacher ratio by town vs `MEDV`, grouped by whether locating along Charles River, it can be seen that higher pupil-teacher ratio by town decreases house prices, whether houses are located along Charles River or not. For the same level of pupil-teacher ratio by town, average `MEDV` seems to be about the same for houses located along Charles River and for houses not located along Charles River.


## 3.12 Proportion of blacks by town
```{r, echo=FALSE, message=FALSE, warning=FALSE}
boston %>%
ggplot() +
geom_point(aes(x = MEDV, y = B, color = factor(CHAS))) +
theme_classic() + 
labs(caption = "Figure 12: Median value of owner-occupied homes for different proportion of blacks by town", y = "1000(Bk - 0.63)^2 where Bk = proportion of blacks by town")
```

Plotting (transformed) proportion of blacks by town vs `MEDV`, grouped by whether locating along Charles River, no obvious pattern can be inferred. Because most data have high (transformed) proportion of blacks by town, and `MEDV` is evenly distributed for houses with high (transformed) proportion of blacks by town, whether houses are located along Charles River or not.

## 3.13 Lower status of the population

LSTAT is calculated to be (proportion of adults without high school education + proportion of male laborers)/2 [@A1]. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
boston %>%
ggplot() +
geom_point(aes(x = MEDV, y = LSTAT, color = factor(CHAS))) +
theme_classic() + 
labs(caption = "Figure 13: Median value of owner-occupied homes for different lower status of the population")
```

Plotting lower status of the population vs `MEDV`, grouped by whether locating along Charles River, it can be seen that higher lower status of the population decreases house prices in a near linear way, whether houses are located along Charles River or not. The negsative linear relationship between lower status of the population and `MEDV` seems to be the same for houses located along Charles River and for houses not located along Charles River.

## 3.14 Correlation plot

```{r, echo=FALSE, message=FALSE, warning=FALSE}
correlations <- cor(boston[,-c(4)], use="everything")
corrplot(correlations, method="circle", type="lower",  sig.level = 0.01, insig = "blank")
```

## 3.15 Prediction model for `MEDV` using linear regression

```{r, echo=FALSE, message=FALSE, warning=FALSE}

boston$CHAS <-as.factor(boston$CHAS)
set.seed(1000)

# 80% of the data as trainning set 
row.number = sample(1:nrow(boston), 0.8*nrow(boston))
train = boston[row.number,]

# 20% of the data as test set 
test = boston[-row.number,]

response <- train$MEDV

lm(MEDV ~ ., data=train) %>%
  tidy() %>%
  knitr::kable(caption = "Coefficient-Level Estimates for a Model Fitted to Estimate Variation in MEDV.",col.names = c("Predictor", "B", "SE", "t", "p"),digits = c(0, 2, 3, 2, 3))

```

As `INDUS` and `AGE` are not significant at 5% significance level, they should be dropped to obtain the new regression model below. 74.52 percent of the variation in the response `MEDV` is explained by new regression model, as its adjusted R-squared is 0.7452.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

mod2 <- lm(MEDV ~ .-INDUS-AGE, data=train) 

summary(mod2) %>%
  tidy() %>%
  knitr::kable(caption = "Coefficient-Level Estimates for a Model Fitted to Estimate Variation in MEDV.",col.names = c("Predictor", "B", "SE", "t", "p"),digits = c(0, 2, 3, 2, 3))
```

```{r,message=FALSE,echo=FALSE, include=FALSE}
# linear regression prediction
lmPredict <- predict.lm(mod2, newdata = test[,-14-7-3])
# calculate RMSE, a measure for model performance
Metrics::rmse(test$MEDV, lmPredict)
```

The calculated RMSE is 5.384884. RMSE is a measure of model performance. Small values indicate good model performance, and large values indicate bad model performance. Because the calculated RMSE for our model is 5.384884, which is small, we think linear regression model is a good one.

## 3.16 Prediction model for `MEDV` using LASSO Regression

```{r, echo=FALSE, message=FALSE, warning=FALSE}
response <- train$MEDV
predictor <- as.matrix(cbind(train$CRIM,train$ZN,train$INDUS,train$CHAS,
                             train$NOX,train$RM,train$AGE,train$DIS,train$RAD,
                             train$TAX,train$PTRATIO,train$B,train$LSTAT))
response <- as.matrix(response) 
laa<- lars(predictor,response,type = 'lasso')
plot(laa)
```

```{r,message=FALSE,echo=FALSE, include=FALSE}
# find the best step size for lasso
optimal_step<- laa$df[which.min(laa$Cp)]

# use the best step size for lasso prediction
newPredict <- predict.lars(laa, newx= as.matrix(cbind(test$CRIM,test$ZN,test$INDUS,test$CHAS,
                             test$NOX,test$RM,test$AGE,test$DIS,test$RAD,
                             test$TAX,test$PTRATIO,test$B,test$LSTAT)), s=optimal_step, type= "fit")

# calculate RMSE, a measure for model performance
rmse(test$MEDV, newPredict$fit)
```

The calculated RMSE is 5.548941 RMSE is a measure of model performance. Small values indicate good model performance, and large values indicate bad model performance. Because the calculated RMSE for our model is 5.548941, which is slightly bigger than the linear model, we think LASSO model is slightly worse than the linear regression model.

# 4 Final result

The best prediction model is given by:

$$MEDV = 39.706131 -0.122058*CRIM + 0.041920*ZN + 2.296506*(CHAS = 1)$$
$$-17.920817*NOX + 3.445127*RM $$
$$-1.506440*DIS+0.306841*RAD$$
$$-0.012808*TAX-0.921847*PTRATIO+0.007965*B-0.549754*LSTAT$$

Therefore, keeping all other variables fixed, houses that are located along Charles River tend to have \$2.296506 thousand higher price than houses that are not located along Charles River. The relationship between whether houses are located along Charles River or not and `MEDV` can be considered as linear. Because the LASSO regression did not produce a better result. However, because `NOX` has an estimated coefficient with magnitude about 8 times the estimated coefficient of `CHAS`, the effect of `NOX` can surpass the effect of locating along Charles River on house prices. People consider air quality to be much more important than all other factors. 

# 5 Appendix

The full Boston housing table:

```{r,message=FALSE,echo=FALSE}
knitr::kable(boston[1:7])
```

```{r,message=FALSE,echo=FALSE}
knitr::kable(boston[8:14])
```

# Reference
